An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-12-14 presentation/background-gpus.tex
updates
final presentation updates

\begin{frame}[t]
\frametitle{Graphics processing units}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Highly parallel and multi-threaded accelerator
    \begin{itemize}
        \item NVIDIA Tesla K20 accelerator: 2496 cores
        \item Focus on \emph{throughput}, not execution speed
    \end{itemize}
\item Spectacular performance for workloads that
    are compute-intensive, data-parallel
    (\emph{fine grained parallelism})
\item Effective use requires careful redesign
    of algorithms, code modification,
    and substantial tuning efforts
\item Vendors: NVIDIA, AMD
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\begin{center}
\includegraphics[width=150px]{img/device-comparison.png}
\centering
\end{center}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Performance considerations}
\begin{itemize}
\item CPU-GPU transfers expensive -
    hide communication cost by keeping GPU busy
\item Expose as much parallelism as possible
    to utilize GPU cores effectively
\item Avoid thread divergence and excessive
    synchronization
\item Memory access:
    \begin{itemize}
        \item Threads must access memory in \emph{coalesced}
            fashion
        \item Threads must do enough work to hide
            memory access latency
    \end{itemize}
\item Strided or uncoalesced memory access patterns:
    use explicitly managed cache called (\emph{shared} memory)
\end{itemize}
\end{frame}

>>>>

\begin{frame}[t]
\frametitle{Graphics processing units}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Highly parallel and multi-threaded accelerator
    \begin{itemize}
        \item NVIDIA Tesla K20 accelerator: 2496 cores
        \item Focus on \emph{throughput}, not execution speed
    \end{itemize}
\item Spectacular performance for workloads that
    are compute-intensive, data-parallel
    (\emph{fine grained parallelism})
\item Effective use requires careful redesign
    of algorithms, code modification,
    and substantial tuning efforts
\item Vendors: NVIDIA, AMD
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\begin{center}
\includegraphics[width=150px]{img/device-comparison.png}
\caption{Design philosophies of CPU and GPU.\newline
Credit: NVIDIA CUDA programming guide}
\centering
\end{center}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Performance considerations}
\begin{itemize}
\item CPU-GPU transfers expensive -
    hide communication cost by keeping GPU busy
\item Expose as much parallelism as possible
    to utilize GPU cores effectively
\item Avoid thread divergence and excessive
    synchronization
\item Memory access:
    \begin{itemize}
        \item Threads must access GPU memory (\emph{global} memory) in \emph{coalesced}
            fashion
        \item Threads must do enough work to hide
            memory access latency
    \end{itemize}
\item Strided or uncoalesced memory access patterns:
    use explicitly managed cache (\emph{shared} memory)
\end{itemize}
\end{frame}

### 2015-12-14 presentation/results.tex
corrections to presentation
final presentation updates

\begin{frame}
\frametitle{Overview of results}
\begin{itemize}
\item Results for tridiagonal solver (NEATO)
\begin{itemize}
    \item Global memory v/s shared memory implementations
    \item Compared against library solvers
        (multicore CPU and GPU)
\end{itemize}
\item Results for compact finite difference application
\begin{itemize}
    \item Profiling results
    \item Scaling results
    \item Comparison with CPU-only approach
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU
\item Upto $2048$ equations in $2048$ variables (2-D)
    and $512^2$ equations in $512$ variables (3-D)
\item GPU: NVIDIA Tesla K20
\end{itemize}
\end{block}

\begin{itemize}
\item For 2-D problems, shared memory clearly better
\item For 3-D problems, shared memory speedup
    is smaller
\item Thread inactivity in shared memory leads
    to reduced performance
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=140px]{img/global-vs-shared-2d.eps}

@@

\begin{frame}{Comparison with library solvers}
\footnotesize
\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU and multicore CPU
\item Upto $2048$ equations in $2048$ variables (2-D)
    and $512^2$ equations in $512$ variables (3-D)
\item GPU: NVIDIA Tesla K20
\item CPU: Intel Xeon E5 v2
\item Do \emph{not} include cost of CPU-GPU transfer
\item Library solver details:
\begin{itemize}
    \footnotesize
    \item Intel MKL \texttt{dgtsv} (Gaussian elimination with partial pivoting)
    \item CUSPARSE \texttt{dgtsv} (Hybrid CR+PCR)
\end{itemize}
\item \texttt{-02} level compiler optimizations
\item Double precision
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - speedup}
Speedups over library implementations:
\begin{center}
\includegraphics[width=180px]{img/bench-2d.eps}
\includegraphics[width=180px]{img/bench-3d.eps}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - timings}
Timings for various solvers:
\begin{table}
\resizebox{\textwidth}{!}{%
\input{table/bench.tex}
}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Note on speedups}
\begin{itemize}
\item Faster than CUSPARSE in almost all cases
\item Comparing against 8 CPU cores,
    speedups over MKL in the range 1-2x
\item Solving the tridiagonal systems
    \emph{least amenable to efficient parallel solution}
\item Objective is to keep the problem on the GPU
\item Multicore CPU performance flattens out
\end{itemize}
\end{frame}

\begin{frame}{Compact finite difference evaluation}
\footnotesize
\begin{block}{Benchmark details}
\begin{itemize}
\item Compact finite difference evaluation in 3-D for a function $f(x, y, z)$
\item Problem size up to $2048^3$ on up to 64 GPUs and 512 CPU cores
\item Tests performed on Palmetto cluster (Infiniband interconnect)
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - profiling}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
\item Most time spent on solving tridiagonal systems
    \begin{itemize}
        \item Justifies optimization effort
    \end{itemize}
\item Significant portion of time spent on
    data permutation
    \begin{itemize}
        \item na\"{\i}ve implementation
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=170px]{img/profiling-1024-64.eps}

\includegraphics[width=170px]{img/profiling-2048-64.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - strong scaling}
\centering
\includegraphics[width=180px]{img/strong-scaling-256.eps}
\includegraphics[width=180px]{img/strong-scaling-512.eps}

Problem sizes: $256^3$ and $512^3$
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - weak scaling}
\centering
\includegraphics[width=180px]{img/weak-scaling-128.eps}
\includegraphics[width=180px]{img/weak-scaling-256.eps}

Problem sizes: $128^3$ and $256^3$ \emph{per process}
\end{frame}

\begin{frame}
\frametitle{Comparison with CPU-only approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Reference implementation: CFDNS
    (Los Alamos National Lab)
\item Uses LU algorithm
\item Parallelized among individual cores
\item Maintain a ratio of 8 CPU cores : 1 GPU for comparison
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{table}
\resizebox{0.75\textwidth}{!}{%
\input{table/compact-refimpl-timings.tex}
}
\end{table}

\includegraphics[width=140px]{img/compact-refimpl-speedups.eps}
\end{column}
\end{columns}
\end{frame}

>>>>

\begin{frame}
\frametitle{Overview of results}
\begin{enumerate}
\item Results for tridiagonal solver (NEATO)
\begin{itemize}
    \item Global memory v/s shared memory implementations
    \item Compared with library solvers
        (multicore CPU and GPU)
\end{itemize}
\item Results for compact finite difference application
\begin{itemize}
    \item Profiling results
    \item Scaling results
    \item Comparison with CPU-only approach
\end{itemize}
\end{enumerate}
\end{frame}

@@

\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU
\item Up to $2048$ systems with $2048$ equations (2-D)
    and $512^2$ systems with $512$ equations (3-D)
\item GPU: NVIDIA Tesla K20
\end{itemize}
\end{block}

\begin{itemize}
\item For 2-D problems, shared memory clearly better
\item For 3-D problems, shared memory speedup
    is smaller
\item \textbf{Thread inactivity} in shared memory leads
    to reduced performance
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=140px]{img/global-vs-shared-2d.eps}

@@

\begin{frame}{Comparison with library solvers}
\footnotesize
\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU and multicore CPU
\item Up to $2048$ equations in $2048$ variables (2-D)
    and $512^2$ equations in $512$ variables (3-D)
\item GPU: NVIDIA Tesla K20
\item CPU: Intel Xeon E5 v2
\item Do \emph{not} include cost of CPU-GPU transfer
\item Library solver details:
\begin{itemize}
    \footnotesize
    \item Intel MKL \texttt{dgtsv} (Gaussian elimination with partial pivoting)
    \item CUSPARSE \texttt{dgtsv} (Hybrid CR+PCR)
\end{itemize}
\item \texttt{-02} level compiler optimizations
\item Double precision
\item Comparison with up to 16 CPU cores
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - speedup}
Speedups over library implementations:

\begin{figure}
\begin{minipage}{.45\textwidth}
    \centering
    \includegraphics[width=\linewidth]{img/bench-2d.eps}
    \caption{2-D problems}
\end{minipage}
\begin{minipage}{.45\textwidth}
    \centering
    \includegraphics[width=\linewidth]{img/bench-3d.eps}
    \caption{3-D problems}
\end{minipage}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Notes on speedups}
\begin{itemize}

\item Faster than CUSPARSE in almost all cases
\begin{itemize}
    \item Can accomodate larger problems than CUSPARSE
\end{itemize}
\item Multicore CPU performance flattens out
\item Compared with 8 CPU cores, speedups over MKL in the range 1-2x
\item Solving the tridiagonal systems
    \emph{least amenable to efficient parallel solution}
\item Primary objective is to keep the problem on the GPU
\end{itemize}
\end{frame}

\begin{frame}{Compact finite difference evaluation}
\begin{block}{Benchmark details}
\begin{itemize}
\item Compact finite difference evaluation in 3-D for a function $f(x, y, z)$
\item Problem size up to $2048^3$ on up to 64 GPUs and 512 CPU cores
\item Tests performed on Palmetto cluster (Infiniband interconnect)
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - profiling}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Most time spent on solving tridiagonal systems
    \begin{itemize}
        \item Justifies optimization effort
    \end{itemize}
\item Compute/communicate ratio increases with
    problem size
\item Significant portion of time spent on
    data permutation
    \begin{itemize}
        \item na\"{\i}ve implementation
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{figure}
\centering
\includegraphics[width=150px]{img/profiling-1024-64.eps}

\includegraphics[width=150px]{img/profiling-2048-64.eps}
\caption{Problem sizes: $1024^3$ (above)
    and $2048^3$ (below) - 64 GPUs}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - strong scaling}
\begin{figure}
\begin{minipage}{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{img/strong-scaling-256.eps}
\caption{Problem size: $256^3$}
\end{minipage}
\begin{minipage}{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{img/strong-scaling-512.eps}
\caption{Problem size: $512^3$}
\end{minipage}
\end{figure}
\centering
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - strong scaling}
\begin{figure}
\begin{minipage}{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{img/weak-scaling-128.eps}
\caption{Problem size \emph{per GPU}: $128^3$}
\end{minipage}
\begin{minipage}{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{img/weak-scaling-256.eps}
\caption{Problem size \emph{per GPU}: $256^3$}
\end{minipage}
\end{figure}
\centering
\end{frame}

\begin{frame}
\frametitle{Comparison with CPU-only approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Reference implementation: CFDNS
    (Los Alamos National Lab)
\item Uses LU algorithm
\item Parallelized among individual cores
\item Maintain a ratio of 8 CPU cores : 1 GPU for comparison
\item Speedups  in the range 3-9x
\item NEATO has strong potential for GPU-based clusters
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{table}
\resizebox{0.75\textwidth}{!}{%
\input{table/compact-refimpl-timings.tex}
}
\end{table}

\includegraphics[width=160px]{img/compact-refimpl-speedups.eps}
\end{column}
\end{columns}
\end{frame}

### 2015-12-14 presentation/tridiagonal-solver.tex
updates
corrections to presentation
final presentation updates

\section{Developing a tridiagonal solver}

@@

\begin{frame}
\frametitle{Thomas algorithm}
\begin{itemize}
\item Derived from Gaussian elimination
\item Requires $2n$ steps and $4n$ storage
\item The most efficient sequential algorithm
\item What about for GPUs?
    \begin{itemize}
        \item inherently sequential
        \item use multiple threads to solve
            individual tridiagonal systems
            (Sakharnykh et al.)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Issues with GPU implementation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Uncoalesced memory accesses
    \item Needs data rearrangement to coalesce accesses
    \item Does not expose enough parallelism
    \item $2n$ steps - can do better on GPU
    \item A case in which \emph{algorithm choice
        significantly impacts GPU performance}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \includegraphics[width=120px]{img/pThomas.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\frametitle{Cyclic reduction}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Buzbee and Goleb (1970)
\item \textbf{Forward reduction} ($log(n)-1$ steps):
    \begin{itemize}
    \item Eliminate odd-indexed equations at each step
    \item Two-by-two system is left
    \end{itemize}
\item \textbf{Backward substitution} ($log(n)-1$ steps):
    \begin{itemize}
    \item Solve for odd-indexed equations using even-indexed values
    \end{itemize}
\item Best case ($n$ parallel threads): requires $2log(n)-1$ steps
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Cyclic reduction}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item $17$ operations per equation,
    compared to $8$ in Thomas algorithm
\item But exhibits more parallelism:
\begin{itemize}
    \item Individual systems can be solved independently
    \item Individual equations can be computed independently
\end{itemize}
\item Equations updated in place: $4n$ storage (same as Thomas algorithm)
\item Synchronization between threads required at each step
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
Forward reduction
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}

@@

\begin{frame}
\frametitle{Issues}
\footnotesize
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Strided memory access:
        stride increases by a factor of 2 at each step
        (decreases during backward substitution)
    \item \textbf{Solution}: use \emph{shared memory} (cache)
    \item Thread activity decreases during forward
        reduction and increases during backward substitution
    \item \textbf{Solution}: schedule threads for each step
        \begin{itemize}
            \footnotesize
            \item conflicts with shared memory usage
        \end{itemize}
    \item Shared memory \emph{bank conflicts}
        due to power-of-2 strides; reduced GPU occupancy
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Diagonals are nearly constant (\emph{near-Toeplitz tridiagonal})
\item Near-Toeplitz trdiagonal matrices occur in other numerical schemes
\begin{itemize}
    \item Alternating direct implicit methods
    \item Line relaxation methods
    \item Poisson solvers
    \item One-dimensional ODEs and PDEs
\end{itemize}
\item Symmetric: but may be broken by boundary conditions
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\scalebox{0.6}{%
\vbox{
\begin{equation*}
\begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
\end{equation*}}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{General approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item {First reduction step: substitute:
    \scalebox{0.8}{
    \vbox{
    \begin{align*}
        a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
        b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
        c_2 &= c_3 = c_4 = \hdots &\equiv c_0
    \end{align*}}}}
\item \textbf{The resulting reduced system is also near-Toeplitz}
\item Each subsequent forward reduction step produces a near-Toeplitz
    tridiagonal system
\begin{itemize}
    \item can be stored compactly
    \item no need to compute for each RHS/time step
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
Forward reduction
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}
\end{column}
\end{columns}
\end{frame}

@@

\textbf{Forward reduction}
\begin{align*}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\end{align*}

@@

where $a^m$, $b^m$ and $c^m$ are the precomputed
coefficients for $i>1$ at the step $m$.
\end{frame}

>>>>

\section{Developing a GPU tridiagonal solver}

@@

\begin{frame}
\frametitle{Thomas algorithm}
\begin{itemize}
\item Derived from Gaussian elimination
\item Requires $2n$ steps and $4n$ storage
\item The most efficient sequential algorithm
\item What about parallelization/GPUs?
    \begin{itemize}
        \item inherently sequential
        \item use multiple threads to solve
            individual tridiagonal systems
            (Sakharnykh et al.)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Issues in GPU implementation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Uncoalesced memory accesses
    \item Needs data rearrangement to coalesce accesses
    \item Does not expose enough parallelism
    \item $2n$ steps - can do better on GPU
    \item A case in which \emph{algorithm choice
        significantly impacts GPU performance}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \includegraphics[width=120px]{img/pThomas.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\frametitle{Cyclic reduction}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Buzbee and Goleb (1970)
\item \textbf{Forward reduction} ($log_2(n)-1$ steps):
    \begin{itemize}
    \item Eliminate odd-indexed equations at each step
    \item Two-by-two system is left
    \end{itemize}
\item \textbf{Backward substitution} ($log_2(n)-1$ steps):
    \begin{itemize}
    \item Solve for odd-indexed equations using even-indexed values
    \end{itemize}
\item Best case ($n$ parallel threads): requires $2log_2(n)-1$ steps
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Cyclic reduction}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item $17n$ operations,
    compared to $8n$ in Thomas algorithm
\item But exhibits more parallelism:
\begin{itemize}
    \item Individual systems can be solved independently
    \item Individual equations can be computed independently
\end{itemize}
\item Equations updated in place: $4n$ storage (same as Thomas algorithm)
\item Synchronization between threads required at each step
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
Forward reduction

(diagonals $a$, $b$, $c$ and RHS $d$)
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}

@@

\begin{frame}
\frametitle{Issues in GPU implementation}
\footnotesize
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Strided memory access:
        stride increases by a factor of 2 at each step
        (decreases during backward substitution)
    \item \textbf{Solution}: use \emph{shared memory} (cache)
    \item Thread activity decreases during forward
        reduction and increases during backward substitution
    \item \textbf{Solution}: schedule threads for each step
        \begin{itemize}
            \footnotesize
            \item conflicts with shared memory usage
        \end{itemize}
    \item Shared memory \emph{bank conflicts}
        due to power-of-2 strides; reduced GPU occupancy
    \item We provide both implementations:
        global memory \emph{and} shared memory
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\frametitle{A novel approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Diagonals are nearly constant (\emph{near-Toeplitz tridiagonal})
\item Symmetry broken by boundary conditions
\item Near-Toeplitz trdiagonal matrices occur in other numerical schemes
\begin{itemize}
    \item Alternating direct implicit methods
    \item Line relaxation methods
    \item Poisson solvers
    \item One-dimensional ODEs and PDEs
\end{itemize}
\item Can be stored compactly compared to general tridiagonal systems
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\scalebox{0.6}{%
\vbox{
\begin{equation*}
\begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
\end{equation*}}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{General approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Each forward reduction step reduces
    a matrix of size $n$ to $n/2$
\item {First reduction step: substitute:
    \scalebox{0.8}{
    \vbox{
    \begin{align*}
        a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
        b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
        c_2 &= c_3 = c_4 = \hdots &\equiv c_0
    \end{align*}}}}
\item \textbf{The resulting reduced system is also near-Toeplitz}
\item Each subsequent forward reduction step produces a near-Toeplitz
    tridiagonal system
\begin{itemize}
    \item can be stored compactly
    \item no need to compute for each RHS/time step
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
Forward reduction
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}
\end{column}
\end{columns}
\end{frame}

@@

\textbf{Forward reduction}

\hspace{3.85cm} \sout{$k_1 = \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}}$}

\hspace{3.85cm} \sout{$a^{\prime}_i = -a_{i-1}k_1$}

\hspace{3.85cm} \sout{$b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2$}

\hspace{3.85cm} \sout{$c^{\prime}_i = -c_{i+1}k_2$} 
\begin{align*}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\end{align*}

@@

where $k_1^m$, $k_2^m$, $a^m$, $b^m$ and $c^m$ are the precomputed
coefficients for $i>1$ at the step $m$.
\end{frame}
